---
title: "Use kableExtra with formattable"
author: "Hao Zhu"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Use kableExtra with formattable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

I know you love `formattable`'s `color_tile` and `color_bar`. I do too. In my mind, these two features are among the best visualization methods in a tabular format. You may wonder if it's possible use it together with `kableExtra` so you can have a nested table with awesome presentation at the same time. Here is how. 

```{r, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(formattable)
library(dplyr)
```

```{r}
mtcars[1:5, 1:4] %>%
  mutate(
    car = row.names(.),
    mpg = color_tile("white", "orange")(mpg),
    cyl = cell_spec(cyl, "html", angle = (1:5)*60, 
                    background = "red", color = "white", align = "center"),
    disp = ifelse(disp > 200,
                  cell_spec(disp, "html", color = "red", bold = T),
                  cell_spec(disp, "html", color = "green", italic = T)),
    hp = color_bar("lightgreen")(hp)
  ) %>%
  select(car, everything()) %>%
  kable("html", escape = F) %>%
  kable_styling("hover", full_width = F) %>%
  column_spec(5, width = "3cm") %>%
  add_header_above(c(" ", "Hello" = 2, "World" = 2))
```

Here is a little bit more explanation. Both `formattable::color_tile` and `formattable::color_bar` returns a function object which can take a numeric vector. That's why you can put things like `(mpg)` after `color_tile("white", "orange")` as you can see in the `mutate` function. This way of using the function may look a little odd but is totally valid (, I think :P).

Also, if you are using it in this way, make sure you put `escape = F` in `kable`. 

On the other hand, `cell_spec()` is a new function in `kableExtra()` to format cells _**before you pipe the table into `kable`**_. Note that you can either specify format (`html` or `latex`) in function or do that via `options(knitr.table.format)` so you don't need to do it everytime. Please refer to the package vignette for details. 

<script>
$(document).ready(function(){
    $('[data-toggle="tooltip"]').tooltip(); 
});
</script>

```{r}
iris[1:10, ] %>%
  mutate(
    Species = cell_spec(Species, color = spec_color(1:10, option = "A"), link = "#",
                        tooltip = paste0("Sepal Length: ", Sepal.Length))
  ) %>%
  mutate_if(is.numeric, function(x){
    cell_spec(x, "html", color = spec_color(x), font_size = spec_font_size(x), bold = T)
  }) %>%
  kable("html", escape = F, align = "c") %>%
  kable_styling("condensed", full_width = F)
```

