---
title: "In-table Plots"
output: 
  html_document:
    theme: cosmo
---

```{r, include=F}
knitr::opts_chunk$set(warning = F, message = F)
```

Sometimes it's a delight to see plots in a table. Here we provide recipes for a few common tasks using both `kableExtra`'s built-in functions and external libraries, including `ggplot2`, `sparkline` and `plotly`.

In terms of `kableExtra`'s built-in plotting functions, since version 1.2.1, `kableExtra` includes these functions for the most common plotting tasks. Since I don't want to introduce an entire `ggplot2` dependency to this package, all these plotting functions were developed using base-R `plot`. Here I need to acknowledge the significant contribution from Bill Evans (@r2evans on github and SO), who wrote the general `spec_plot` function and improved the coding structure of these `spec_***` plotting functions. 

```{r}
library(kableExtra)
library(plotly)
library(sparkline)
library(tidyverse)
```

## Summary Stats Table

```{r}
  
```

### kableExtra built-in

```{r}

```


